---
title: "data-exploration"
output: html_document
---

```{r, load data}
# library(HDSinRdata)
# library(readr)
# library(dplyr)
# library(stringr)

library(haven)
library(tidyverse)



# Data is from: https://wwwn.cdc.gov/nchs/nhanes/continuousnhanes/default.aspx?Cycle=2021-2023
# Description of data: https://wwwn.cdc.gov/nchs/nhanes/search/datapage.aspx?Component=Questionnaire&Cycle=2021-2023

data_alq = read_xpt("./data/ALQ_L.xpt")     # Alcohol
data_bmx = read_xpt("./data/BMX_L.xpt")     # Body measurements
data_bpq = read_xpt("./data/BPQ_L.xpt")     # Blood Pressure & Cholesterol
data_demo = read_xpt("./data/DEMO_L.xpt")   # Demographics data
data_fsq = read_xpt("./data/FSQ_L.xpt")   # Not related to health demographics and not used.
data_inq = read_xpt("./data/INQ_L.xpt")     # Income information
data_paq = read_xpt("./data/PAQ_L.xpt")     # Physical activity adult
data_paqy = read_xpt("./data/PAQY_L.xpt")   # Physical activity youth
data_smq = read_xpt("./data/SMQ_L.xpt")     # Smoke usage
data_smqr = read_xpt("./data/SMQRTU_L.xpt") # Smoking recent


# Join the data into a single df
data_total = left_join(data_alq, data_bmx)
data_total = left_join(data_total, data_bpq)
data_total = left_join(data_total, data_demo)
# data_total = left_join(data_total, data_fsq)
data_total = left_join(data_total, data_inq)
data_total = left_join(data_total, data_paq)
data_total = left_join(data_total, data_paqy)
data_total = left_join(data_total, data_smq)
data_total = left_join(data_total, data_smqr)


```



**Questions**
How are age, sex, race/ethnicity, physical activity, and smoking associated with BMI?
  - SLR, MLR
Can we predict whether an adult is obese using demographic and lifestyle predictors?
  - Use selection method, and PCA.
Does alcohol consumption remain associated with BMI after adjusting smoking status and physical activity?
  - chi squared (todo: research this topic) Hypothesis testing. 




```{r, q1}

# How are age, gender, race/ethnicity, physical activity, and smoking associated with BMI?




# Linear regression: BMI ~ age + gender + physical activity + smoking

model_df = data_total %>%
  transmute(
    BMI = BMXBMI,
    age = RIDAGEYR,
    gender = RIAGENDR,
    days_physical = PAD800,
    smoking = SMQ040
  ) %>%
  # Remove special missing/refused codes
  filter(
    !is.na(BMI),
    !is.na(age),
    !is.na(gender),
    !is.na(days_physical),
    !is.na(smoking),
    !days_physical %in% c(7777, 9999), # NHANES codes: 7777-Refused, 9999-Don't know
    smoking %in% c(1, 2, 3)  # NHANES codes: 1-Every day, 2-Some days, 3-Not at all
  ) %>%
  mutate(
    gender_value = factor(gender, levels = c(1, 2), labels = c("Male", "Female")),
    smoking = factor(smoking)  # keeps smoking as categorical
  )


# Fit model
fit_bmi = lm(BMI ~ age + gender + days_physical + smoking, data = model_df)

# Results
summary(fit_bmi)

# Optional diagnostics
par(mfrow = c(2, 2)) # divide the plotting area into 2x2 grid
plot(fit_bmi)


# Table of coefficients
coef_table = summary(fit_bmi)$coefficients
coef_table

# Non significant coefficients
nonsig = coef_table[coef_table[, "Pr(>|t|)"] > 0.05, , drop = FALSE]
nonsig


```




```{r q1 subset selection}

# install.packages("leaps")  # run once
library(leaps)

# Build matrix from same formula
x = model.matrix(BMI ~ age + gender + days_physical + smoking, data = model_df)[, -1]
y = model_df$BMI

subset_fit = regsubsets(x = x, y = y, nvmax = ncol(x), method = "exhaustive")
subset_sum = summary(subset_fit)

# Compare models by adjusted R^2, Cp, BIC
subset_sum$adjr2
subset_sum$cp
subset_sum$bic

# Best model size by each criterion
best_adjr2_size = which.max(subset_sum$adjr2)
best_cp_size    = which.min(subset_sum$cp)
best_bic_size   = which.min(subset_sum$bic)

best_adjr2_size
best_cp_size
best_bic_size

# Coefficients for best BIC, cp, adjR^2 model
coef(subset_fit, best_bic_size)
coef(subset_fit, best_cp_size)
coef(subset_fit, best_adjr2_size)


# Refit with selected terms
final_fit = lm(BMI ~ age + days_physical + smoking, data = model_df)
summary(final_fit)

```






```{r, q2}

# Can we predict whether an adult is obese using demographic and lifestyle predictors?
#   - Use selection method, and PCA.

# We will need to clean the data here, looking for only predictors that relate to BMI.
# Idea is to use subset selection or PCA to find the most important predictors.

# Relate blood pressure, bmi, vs other demographic measures.


# https://github.com/ryantibs/best-subset
library(devtools)
install_github(repo="ryantibs/best-subset", subdir="bestsubset")




```




```{r, q3}

# Does alcohol consumption remain associated with BMI after adjusting smoking status and physical activity?
#   - chi squared (todo: research this topic) Hypothesis testing.
#Is alcohol use category (ALQ_L) associated with high blood pressure (BPQ_L — "ever told you had high blood pressure")?
library(haven)
library(tidyverse)
library(gmodels)
library(corrplot)


cst = data_demo %>%      #creating the dataframe for the chi square goodness of fit test
  inner_join(data_alq, by = "SEQN") %>%
  inner_join(data_bpq, by = "SEQN") %>%
  filter(RIDAGEYR >= 18)

cst$ALQ111[is.na(cst$ALQ111)] = 0 # replacing na values with 0 to fit testing
cst$ALQ130[is.na(cst$ALQ130)] = 0


cst["HighBP"]= c( " ")  #creating column space for high blood pressure
cst["Drinks"]= c( " ")  #column for amount of drinks had per day



cst_clean = cst%>% mutate(cst, Drinks = if_else(ALQ111 == 2, "Non-drinker",
                                         if_else(ALQ111 ==1 & ALQ130 <=2,"Light",
                                                 if_else(ALQ111 == 1 & ALQ130 <= 4, "Moderate",
                                                         if_else(ALQ111 ==1 & ALQ130 > 4, "High",
                                                                 if_else(ALQ111 == 9, NA,
                                                                         if_else(ALQ111 == ".", NA, 
                                                                                 if_else(ALQ111==1 & ALQ130 ==777, NA,
                                                                                         if_else(ALQ111==1 & ALQ130 ==999, NA,
                                                                                                 if_else(ALQ111==1 & ALQ130 == ".",NA,"Non-drinker"))))))))),
                     HighBP = if_else(BPQ080 == 1, "Yes", "No"),

      HighBP  = factor(HighBP,  levels = c("No", "Yes")),
      Drinks   = factor(Drinks,
                       levels = c("Non-drinker", "Light",
                                  "Moderate", "High")))

cst_clean= cst_clean[!is.na(cst_clean$Drinks),]

test_data = cst_clean %>% select(HighBP, Drinks)
contingency_table = table(test_data$HighBP, test_data$Drinks)
print(contingency_table)

chisq_res = chisq.test(contingency_table)  # test of independence
chisq_res
chisq_res$expected
print(round(chisq_res$expected,2))
print(round(chisq_res$residuals, 2)) 

corrplot(chisq_res$residuals, is.cor = FALSE)


```







```{r}
#Q4 - Can we predict whether an individual or household is food secure or insecure based on the following variables

library(haven)       
library(tidyverse)
library(caret)       
library(rpart)       
library(rpart.plot)

data_fsq$FSDAD[is.na(data_fsq$FSDAD)] = 0

data_fsq["Food_insecure"]=c(" ") #column that says whether or not the individual is classified as food insecure or not
data_fsq["On_Snap"]=c(" ")  #column that says whether or not they were receiving SNAP benefits in the last 12 months
data_fsq["EFood"]= c(" ")  #column that says whether or not they received emergency food
data_fsq["WIC"]=c(" ")   #column that says whether or not the household received WIC program benefits
data_fsq["NSnap"]= c(" ") #column that says the amount of people in a household received SNAP benefits

# adding N/A as a value to clean later on
fsq_test = data_fsq %>% mutate(data_fsq, Food_insecure= if_else(FSDAD>=3,1, #1 = yes, 2= no
                                                      if_else(FSDAD>0 & FSDAD <3, 2,
                                                              if_else(FSDAD == ".", NA, 2))),
                          On_Snap = if_else(FSQ012>2, NA,
                                                  if_else(FSQ012 == ".",NA, FSQ012)),
                  EFood = if_else(FSD151>2, NA,
                                        if_else(FSD151 == ".",NA,FSD151)),
                  WIC = if_else(FSD162>2, NA,
                                if_else(FSD162== ".", NA, FSD162)),
                  NSnap =if_else(FSD165N>7, NA,
                                 if_else(FSD165N==".",NA,FSD165N)))


demo_test = data_demo %>%
  select(SEQN, age=  RIDAGEYR, sex = RIAGENDR, race = RIDRETH3, education = DMDEDUC2, IPR = INDFMPIR, hh_size = DMDHHSIZ)

test_data2 = fsq_test %>%
    inner_join(demo_test, by = "SEQN") %>%
        select(Food_insecure,On_Snap,EFood,WIC,NSnap,age, sex, race, education, IPR, hh_size)

test_data2 = test_data2 %>% mutate(
                    Food_insecure = factor(Food_insecure, levels = c(1,2), labels = c("Insecure","Secure")),
                    sex = factor(sex, levels = c(1,2), labels = c("Male","Female")),
                    race = factor(race),
                    education = factor(education),
                    On_Snap = factor(On_Snap, levels = c(1,2), labels = c("Yes","No")),
                    EFood= factor(EFood,levels = c(1,2), labels=c("Yes","No")),
                    WIC = factor(WIC,levels = c(1,2), labels = c("Yes","No"))) 
test_data2 = test_data2 %>% na.omit(test_data2)

set.seed(1)



trainsize = floor(0.70 * nrow(test_data2))
indices = sample(seq_len(nrow(test_data2)), size = trainsize)
tdata = test_data2[-indices,]
trdata = test_data2[indices,]

prop.table(table(trdata$Food_insecure))

FoodTree = rpart(Food_insecure~ On_Snap + EFood+ WIC + IPR + hh_size + age + sex + race + education,
                 trdata, method = "class", na.action = na.rpart, control = rpart.control(
                   cp = 0.005,
                   minsplit = 20,
                   maxdepth = 6
                 ))

printcp(FoodTree)
bestv = FoodTree$cptable[which.min(FoodTree$cptable[, "xerror"]),"CP"]
prunetree = prune(FoodTree,cp=bestv)
rpart.plot(prunetree,
           type = 4,
           extra = 104,
           fallen.leaves = TRUE)

tree_class = predict(prunetree, tdata, type="class")
treeprob = predict(prunetree, tdata, type = "prob")[, "Insecure"]

cm = confusionMatrix(tree_class, tdata$Food_insecure, positive = "Insecure")
print(cm)
FN = cm$table[2,1]  
TP = cm$table[1,1]    
FNR = FN / (FN + TP)



```


