---
title: "data-exploration"
output: html_document
---

```{r, load data}
# library(HDSinRdata)
# library(readr)
# library(dplyr)
# library(stringr)

library(haven)
library(tidyverse)



# Data is from: https://wwwn.cdc.gov/nchs/nhanes/continuousnhanes/default.aspx?Cycle=2021-2023
# Description of data: https://wwwn.cdc.gov/nchs/nhanes/search/datapage.aspx?Component=Questionnaire&Cycle=2021-2023

data_alq = read_xpt("./data/ALQ_L.xpt") # Alcohol
data_bmx = read_xpt("./data/BMX_L.xpt") # Body measurements
data_bpq = read_xpt("./data/BPQ_L.xpt") # Blood Pressure & Cholesterol
data_demo = read_xpt("./data/DEMO_L.xpt") # demographics data
# data_fsq = read_xpt("./data/FSQ_L.xpt") # Not related to health demographics and not used.
data_inq = read_xpt("./data/INQ_L.xpt") # Income information

data_smqr = read_xpt("./data/SMQRTU_L.xpt") # smoking recent
data_smq = read_xpt("./data/SMQ_L.xpt") # smoke usage
data_paq = read_xpt("./data/PAQ_L.xpt") # physical activity adult
data_paqy = read_xpt("./data/PAQY_L.xpt") # physical activity youth

data_total = left_join(data_alq, data_bmx)
data_total = left_join(data_total, data_demo)
data_total = left_join(data_total, data_inq)
# data_total = left_join(data_total, data_fsq)
data_total = left_join(data_total, data_smqr)
data_total = left_join(data_total, data_smq)
data_total = left_join(data_total, data_paq)
data_total = left_join(data_total, data_paqy)


glimpse(data_total)


## TODO: Create a function that returns the column name from a df, or manually
# Get the label for a column named 'USUBJID'
column_description <- attr(data_total$ALQ111, "label")
print(column_description)






#attr# summary(is.na(data_total))






```



**Questions**
How are age, sex, race/ethnicity, physical activity, and smoking associated with BMI?
  - SLR, MLR
Can we predict whether an adult is obese using demographic and lifestyle predictors?
  - Use selection method, and PCA.
Does alcohol consumption remain associated with BMI after adjusting smoking status and physical activity?
  - chi squared (todo: research this topic) Hypothesis testing. 




```{r, q1}

# How are age, sex, race/ethnicity, physical activity, and smoking associated with BMI?
#   - SLR, MLR

#race is a complex range of values, categories so we will leave it out for now; can include this later

age_lab = attr(data_total$RIDAGEYR, "label")

age_value = data_total$RIDAGEYR
sex = data_total$RIAGENDR

# Create a range based on the data fields: 
# Drop 999 PAD800
physical_activity = 


# BMI is found in the data_bmx file - data_total$
bmi_measure = data_total$BMXBMI



```



```{r, q2}

# Can we predict whether an adult is obese using demographic and lifestyle predictors?
#   - Use selection method, and PCA.

# We will need to clean the data here, looking for only predictors that relate to BMI.
# Idea is to use subset selection or PCA to find the most important predictors.

# Relate blood pressure, bmi, vs other demographic measures.

```




```{r, q3}

# Does alcohol consumption remain associated with BMI after adjusting smoking status and physical activity?
#   - chi squared (todo: research this topic) Hypothesis testing.


```

