---
title: "data-exploration"
output: html_document
---

```{r, load data}
# library(HDSinRdata)
# library(readr)
# library(dplyr)
# library(stringr)

library(haven)
library(tidyverse)



# Data is from: https://wwwn.cdc.gov/nchs/nhanes/continuousnhanes/default.aspx?Cycle=2021-2023
# Description of data: https://wwwn.cdc.gov/nchs/nhanes/search/datapage.aspx?Component=Questionnaire&Cycle=2021-2023

data_alq = read_xpt("./data/ALQ_L.xpt")     # Alcohol
data_bmx = read_xpt("./data/BMX_L.xpt")     # Body measurements
data_bpq = read_xpt("./data/BPQ_L.xpt")     # Blood Pressure & Cholesterol
data_demo = read_xpt("./data/DEMO_L.xpt")   # Demographics data
# data_fsq = read_xpt("./data/FSQ_L.xpt")   # Not related to health demographics and not used.
data_inq = read_xpt("./data/INQ_L.xpt")     # Income information
data_paq = read_xpt("./data/PAQ_L.xpt")     # Physical activity adult
data_paqy = read_xpt("./data/PAQY_L.xpt")   # Physical activity youth
data_smq = read_xpt("./data/SMQ_L.xpt")     # Smoke usage
data_smqr = read_xpt("./data/SMQRTU_L.xpt") # Smoking recent


# Join the data into a single df
data_total = left_join(data_alq, data_bmx)
data_total = left_join(data_total, data_bpq)
data_total = left_join(data_total, data_demo)
# data_total = left_join(data_total, data_fsq)
data_total = left_join(data_total, data_inq)
data_total = left_join(data_total, data_paq)
data_total = left_join(data_total, data_paqy)
data_total = left_join(data_total, data_smq)
data_total = left_join(data_total, data_smqr)


```



**Questions**
How are age, sex, race/ethnicity, physical activity, and smoking associated with BMI?
  - SLR, MLR
Can we predict whether an adult is obese using demographic and lifestyle predictors?
  - Use selection method, and PCA.
Does alcohol consumption remain associated with BMI after adjusting smoking status and physical activity?
  - chi squared (todo: research this topic) Hypothesis testing. 




```{r, q1}

# How are age, gender, race/ethnicity, physical activity, and smoking associated with BMI?




# Linear regression: BMI ~ age + gender + physical activity + smoking

model_df = data_total %>%
  transmute(
    BMI = BMXBMI,
    age = RIDAGEYR,
    gender = RIAGENDR,
    days_physical = PAD800,
    smoking = SMQ040
  ) %>%
  # Remove special missing/refused codes
  filter(
    !is.na(BMI),
    !is.na(age),
    !is.na(gender),
    !is.na(days_physical),
    !is.na(smoking),
    !days_physical %in% c(7777, 9999), # NHANES codes: 7777-Refused, 9999-Don't know
    smoking %in% c(1, 2, 3)  # NHANES codes: 1-Every day, 2-Some days, 3-Not at all
  ) %>%
  mutate(
    gender_value = factor(gender, levels = c(1, 2), labels = c("Male", "Female")),
    smoking = factor(smoking)  # keeps smoking as categorical
  )


# Fit model
fit_bmi = lm(BMI ~ age + gender + days_physical + smoking, data = model_df)

# Results
summary(fit_bmi)

# Optional diagnostics
par(mfrow = c(2, 2)) # divide the plotting area into 2x2 grid
plot(fit_bmi)


# Table of coefficients
coef_table = summary(fit_bmi)$coefficients
coef_table

# Non significant coefficients
nonsig = coef_table[coef_table[, "Pr(>|t|)"] > 0.05, , drop = FALSE]
nonsig


```




```{r q1 subset selection}

# install.packages("leaps")  # run once
library(leaps)

# Build matrix from same formula
x = model.matrix(BMI ~ age + gender + days_physical + smoking, data = model_df)[, -1]
y = model_df$BMI

subset_fit = regsubsets(x = x, y = y, nvmax = ncol(x), method = "exhaustive")
subset_sum = summary(subset_fit)

# Compare models by adjusted R^2, Cp, BIC
subset_sum$adjr2
subset_sum$cp
subset_sum$bic

# Best model size by each criterion
best_adjr2_size = which.max(subset_sum$adjr2)
best_cp_size    = which.min(subset_sum$cp)
best_bic_size   = which.min(subset_sum$bic)

best_adjr2_size
best_cp_size
best_bic_size

# Coefficients for best BIC, cp, adjR^2 model
coef(subset_fit, best_bic_size)
coef(subset_fit, best_cp_size)
coef(subset_fit, best_adjr2_size)


# Refit with selected terms
final_fit = lm(BMI ~ age + days_physical + smoking, data = model_df)
summary(final_fit)

```






```{r, q2}

# Can we predict whether an adult is obese using demographic and lifestyle predictors?
#   - Use selection method, and PCA.

# We will need to clean the data here, looking for only predictors that relate to BMI.
# Idea is to use subset selection or PCA to find the most important predictors.

# Relate blood pressure, bmi, vs other demographic measures.


# https://github.com/ryantibs/best-subset
library(devtools)
install_github(repo="ryantibs/best-subset", subdir="bestsubset")




```




```{r, q3}

# Does alcohol consumption remain associated with BMI after adjusting smoking status and physical activity?
#   - chi squared (todo: research this topic) Hypothesis testing.


```

